\documentclass[runningheads,a4paper]{llncs}
\usepackage{amsmath,amssymb,amsfonts,mathrsfs,stmaryrd}
\usepackage{bussproofs}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{algorithm2e}
\usepackage{wrapfig}
\usepackage{url, listings, color}
%\usepackage{hyperref}

%two column float page must be 90% full
\renewcommand\dblfloatpagefraction{.99}
%two column top float can cover up to 80% of page
\renewcommand\dbltopfraction{.99}
%float page must be 90% full
\renewcommand\floatpagefraction{.99}
%top float can cover up to 80% of page
\renewcommand\topfraction{.99}
%bottom float can cover up to 80% of page
\renewcommand\bottomfraction{.99}
%at least 10% of a normal page must contain text
\renewcommand\textfraction{.01}

\newcommand{\code}[1] {{\footnotesize\sffamily #1}}
\newcommand{\transml}[1]{{\textit{trans}ML\ }}

\titlerunning{Engineering Bidirectional Transformations}
\authorrunning{Richard F. Paige}

\begin{document}

\title{Engineering Bidirectional Transformations}

\author{
Richard F. Paige, richard.paige@york.ac.uk}
%
\institute{
Department of Computer Science, University of York, United Kingdom
}

\maketitle

\begin{abstract}
Bidirectional transformations, like software, need to be carefully engineered in order to provide guarantees about their correctness, completeness, acceptability and usability. This paper summarises a collection of lectures pertaining to engineering bidirectional transformations using Model-Driven Engineering techniques and technologies. It focuses on stages of a typical engineering lifecycle, starting with requirements and progressing to implementation and verification. It summarises Model-Driven Engineering approaches to capturing requirements, architectures and designs for bidirectional transformations, and suggests an approach for verification as well. It concludes by describing some challenges for future research into engineering bidirectional transformations.
\end{abstract}

\input{intro}

\input{state-of-the-art}

\input{requirements}

\input{architecture-and-design}

\input{verification}

\input{conclusions}

\vspace{4mm}

\noindent\textbf{Acknowledgements} Parts of this work were supported by the European Commission's 7th Framework Programme, through grant \#611125 (MONDO). The author also acknowledges the support of Innovate UK and the Aerospace Technology via the SECT-AIR grant, and the EPSRC, for their support for the Summer School in Bidirectional Transformations. The author thanks Dimitris Kolovos, Chris Poskitt, Arend Rensink, Mike Dodds, Esther Guerra and Juan de Lara for many useful discussions and collaboration on the topics presented in this paper, and to the reviewers of this manuscript for the helpful suggestions and advice.

\bibliographystyle{unsrt}
\bibliography{bibliography,references-amt}  % sigproc.bib is the name of the Bibliography 

\end{document}
